```{r,message = FALSE,echo = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(knitcitations)
library(kableExtra)
opts_chunk$set(fig.width = 10,
               eval = FALSE,         # Evaluate code chunks
               fig.height = 5,
               cache = FALSE)
cite_options(citation_format = "pandoc", max.names = 3, style = "html", hyperlink = "to.html")
```

---
title: 'Long non-coding RNA Discovery'
subtitle: '*Steven Roberts* and  *Zach Bengston*'
bibliography: DRAFT_lncRNA/DRAFT_lncRNA.bib
---

Initial publication year: TBD <br> [How to cite](https://marineomics.github.io/#How_to_Cite)

# Introduction



## Notes on running code. This workflow is written such that a user an fully reproduce the code. 



# Downloading data

The data used for this workflow is from Acropora pulcra, and part of the E5 (https://e5coral.org) project. see https://github.com/urol-e5/deep-dive for more details.

```{r, engine='bash'}
mkdir ../data/lncRNA

wget -r \
--no-directories --no-parent \
-P ../data/lncRNA \
-A "*fastq.gz" https://gannet.fish.washington.edu/Atumefaciens/20230519-E5_coral-fastqc-fastp-multiqc-RNAseq/A_pulchra/trimmed/

```




# References {#refs}
