---
layout: default
title: "Panel Seminars"
nav_order: 1
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# About

The MarineOmics working group is part of the [RCN for Evolution in Changing Seas](https://rcn-ecs.github.io/), and was formed to promote reproducible and robust genomic research in marine science. Methods in next generation sequencing and bioinformatics advance at a pace that quickly render any static set of best practices as obsolete. To support robust 'omics research now and moving forward, the working group will disseminate guidelines through an easily-interfaced and dynamic website that will be deployed in August 2021. *Keep an eye on this URL for updates!*

# Panel Seminar Series

To facilitate discussions about practices and pitfalls in applying 'omics to non-model systems, we are hosting Q & A panel seminars with experts on topics ranging from RADseq, population genomics with WGS, and transcriptomics. Seminars are every two weeks, on **Fridays from 11:00-12:15 US Central Time**. 3-5 expert panelists will discuss each topic, with questions coming from the working group and the audience. If you would like to submit a question for the panelists ahead of time, please send an email to rcnecs.marineomics[at]gmail.com . 

Everyone is welcome to join! All seminars will be recorded and available on this website.    

**To receive a Zoom link, please fill out the registration form [here](
https://auburn.zoom.us/meeting/register/tZ0ode2rpz4sG9Il_TuYexWfqd6HyjdQ58Zi).**  

## Seminar Schedule

| Date        | Topic       | Panelists     | Video     |
| :---        | :---        | :---          | :---      |
| May 7         | GBS/RADseq genotyping pipelines     | [Melanie LaCava](https://melanielacava.weebly.com/), [Alicia Mastretta-Yanes](http://mastrettayanes-lab.org/), [Isaac Overcast](https://isaacovercast.github.io/), [Jon Puritz](https://www.marineevoeco.com/)   | [View video](https://www.youtube.com/watch?v=C74GBESeIq4)     |
| May 21       |  WGS genotyping (low and high coverage)     | [Jeffrey Good](http://www.thegoodlab.org/), [Mikhail "Misha" Matz](https://matzlab.weebly.com/), [Noah Reid](http://nreid.github.io/), [Tim Sackton](https://scholar.harvard.edu/tsackton), [Nina Overgaard Therkildsen](https://www.therkildsenlab.com/) | [View Video](https://www.youtube.com/watch?v=ObJ3OT3dwSw)    |
| June 4       |  Population genomic analyses and filtering data     | [Zach Gompert](https://gompertlab.com/), [Morgan Kelly](http://www.morgankelly.biology.lsu.edu/), [Katie Lotterhos](https://sites.google.com/site/katielotterhos/home), [Jeff Ross-Ibarra](https://rilab.ucdavis.edu/) |  [View Video](https://youtu.be/dZyuqVEGPMo)   |
| June 18       |  RNAseq (experimental design through differential expression)     | [Ana Conesa](http://conesalab.org/), [Sarah Davies](http://sites.bu.edu/davieslab/), [Martin Hölzer](https://hoelzer-lab.github.io/), [Joanna Kelley](https://labs.wsu.edu/genomes/), [Harold Pimentel](https://pimentellab.com/)     |     |

# Reproducibility in Marine Genomics  
The working group is also currently hosting three undergraduate research interns with funding from the [RCN-ECS](https://rcn-ecs.github.io/). These interns are part of a sub-committee conducting a literature survey to characterize trends in data availability, code accessibility, and methods reproducibility in papers that use genomics to study adaptation in marine systems. The results of this survey will be shared in a peer-reviewed article, along with guidance to improve the state of reproducibility in the field. 

# Working Group Members  

#### [Katherine Silliman, PhD](https://ksilliman.weebly.com/), Auburn University - kes0132[at]auburn.edu
#### [Sam Bogan](http://www.hofmannlab.com/sam-bogan.html), UC-Santa Barbara -snbogan[at]ucsb.edu
#### [Alan Downey-Wall](https://cos.northeastern.edu/people/alan-downey-wall/), Northeastern University 
#### Chris Mantegna, University of Washington  
#### [Danielle Davenport](https://www.researchgate.net/profile/Danielle-Davenport-2), University of Queensland  
#### [Hanny Rivera, PhD](https://www2.whoi.edu/staff/hrivera/), Woods Hole Oceanographic Institute  
#### Jecy Klinkam, College of William and Mary  
#### [Joanna Griffiths, PhD](https://joannasgriffiths.wordpress.com/), UC-Davis
#### [Joe McGirr, PhD](https://joemcgirr.github.io/), UC-Davis 
#### Kaitlin Macaranas, CSU-Bakersfield 
#### Megan Guidry, University of Rhode Island
#### Natalie Ameral, University of Rhode Island & RI Dept. of Environmental Management  
#### Nicolas Lou, Cornell University
#### [Sara Schaal](https://schaals.wixsite.com/smschaal), Northeastern University 
#### [Sara Wuitchik, PhD](https://sjswuitchik.weebly.com/), Harvard University


### Faculty Advisors
#### [Andrew Whitehead, PhD](https://whiteheadresearch.wordpress.com/), UC-Davis 
#### [Jon Puritz, PhD](https://www.marineevoeco.com/), University of Rhode Island 
#### [Steven Roberts, PhD](http://faculty.washington.edu/sr320/?page_id=10850), University of Washington



# Contact 
For more information about the working group or panel seminars, contact Katherine Silliman: kes0132[at]auburn.edu  
For more information about the reproducibility review and undergraduate internship, contact Sam Bogan: snbogan[at]ucsb.edu



``````{r, echo=FALSE,eval=FALSE}
library(knitr)
my.jekyll.site <- "marineomics.github.io"
KnitPost <- function(input, base.url = my.jekyll.site) {
  opts_knit$set(base.url = base.url)
  fig.path <- paste0("images/", sub(".Rmd$", "", basename(input)), "/")
  opts_chunk$set(fig.path = fig.path)
  opts_chunk$set(fig.cap = "center")
  render_jekyll()
  knit(input, envir = parent.frame())
}

setwd("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/")
KnitPost("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/index.Rmd")

```
