---
layout: default
title: "About"
nav_order: 1
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Rigorous, Reproducible Research in Marine Genomics


Welcome! We are a working group dedicated to promoting reproducible and robust genomic research in marine science and evolutionary biology.

## Organizers:

Katherine Silliman and Sam Bogan

## Faculty Mentors

Jon Puritz and Andrew Whitehead

## Working group members





``````{r, echo=FALSE,eval=FALSE}
library(knitr)
my.jekyll.site <- "marineomics.github.io/Template_Site_1/"
KnitPost <- function(input, base.url = my.jekyll.site) {
  opts_knit$set(base.url = base.url)
  fig.path <- paste0("images/", sub(".Rmd$", "", basename(input)), "/")
  opts_chunk$set(fig.path = fig.path)
  opts_chunk$set(fig.cap = "center")
  render_jekyll()
  knit(input, envir = parent.frame())
}

setwd("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/about/")
KnitPost("C:/Users/jmcgirr/Documents/GitHub/MarineOmics/Template_Site_1.github.io/about/about.Rmd")

```
